---
title: "Spatial taphonomy of the Late Villafranchian vertebrate locality of Tsiotra Vryssi, Mygdonia Basin, Greece"
subtitle: "Fabric analysis"
author: "Domenico Giusti"
bibliography: "../doc/manuscript/tsr_rr.bib"
output: 
  html_document: 
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, eval=TRUE, echo=FALSE, message=FALSE}
# load required packages
library(circular)
library(CircStats)
library(RFOC)
library(plotrix)
library(heR.Misc)
```

```{r, eval=TRUE, echo=FALSE, message=FALSE}
getwd()
```

## Reading and subsetting data

```{r, eval=TRUE, echo=FALSE, message=FALSE, results='hide'}
# reading data
summary(X.df)

# remove plunge/bearing NA's
X.circ <- X.df[-which(is.na(X.df$marks.bearing)),]
X.circ <- X.circ[-which(is.na(X.circ$marks.plunge)),]

# subset articulated/isolated bones
X.circ <- X.circ[which(X.circ$marks.articulation == "isolated"),]
#X.circ.art <- X.circ[which(X.circ$articulation == "articulated"),]

# remove length NA's & subset length > 20mm
X.circ <- X.circ[-which(is.na(X.circ$marks.length.x)),]
X.circ <- X.circ[which(X.circ$marks.length.x >= 20),]

# subset excavation year (2015-2016)
#library(dplyr)
#X <- dplyr::filter(X, date < as.Date("2017-01-01"))
#summary(X)

# remove turtle shell
X.circ$marks.description <- factor(X.circ$marks.description)
summary(X.circ$marks.description)
X.circ <- X.circ[which(X.circ$marks.class %in% c("Mammalia",NA)),]
#X.circ <- X.circ[-which(X.circ$description %in% c("?turtle shell","shell")),]

# remove teeth
X.circ$marks.type.y <- factor(X.circ$marks.type.y)
summary(X.circ$marks.type.y)
X.circ <- X.circ[-which(X.circ$marks.type.y == "Tooth"),]

summary(X.circ)
```

## Global Circular statistics

### Rose diagrams & Kernel Density Estimates

```{r}
# circular
bearing.circ <- circular(X.circ$marks.bearing, type="angles", units="degrees", template="geographics")
plot(bearing.circ, cex=1, bin=36, stack=TRUE, sep=0.05, shrink=1.3)
circular::rose.diag(bearing.circ, bins=12, col="darkgrey", prop=1.5, add=TRUE)
lines(circular::density.circular(bearing.circ, bw=bw.cv.ml.circular(bearing.circ)))
arrows.circular(mean(bearing.circ))
```

```{r, message=FALSE, warning=FALSE, results='hide'}
# heR.Misc
heR.Misc::rose(X.circ$marks.bearing, bins=36, col="red", rings=TRUE, labels=TRUE, rscale=2)
heR.Misc::rose(X.circ$marks.bearing, bins=12, col="red", rings=TRUE, labels=TRUE, rscale=2)
heR.Misc::rose2(X.circ$marks.bearing, nplumes=36, shrink=2, bg="white", mag.col="red", col.ring="grey", ring.labels=FALSE, cex.lab=1, cex.dir=1, main="Fig.1c")
```

### Two-dimensional summary statistics

```{r}
# convert degrees to radians
plunge <- rad(X.circ$marks.plunge)
bearing <- rad(X.circ$marks.bearing)

# summary
deg(circ.summary(bearing))
deg(circ.summary(plunge))
# mean
deg(circ.mean(bearing))
deg(circ.mean(plunge))
# dispersion
deg(circ.disp(bearing))
deg(circ.disp(plunge))
```

### Uniformity tests

```{r}
## Rayleigh test for unimodal departure from uniformity
### tests the sample mean resultant length for uniformity against von Mises alternatives
circular::rayleigh.test(bearing.circ)
r.test(bearing)

## Kuiper test
### omnibus test used to detect multimodal departures from uniformity
circular::kuiper.test(bearing.circ)
kuiper(bearing)

## Watson test
circular::watson.test(bearing.circ)
watson(bearing, dist = "uniform")
watson(bearing, dist = "vm")

## rao's spacing test
circular::rao.spacing.test(bearing.circ)
rao.spacing(bearing)
```

### Equal area (Schmidt) diagrams

```{r, message=FALSE, warning=FALSE}
# equal area (schmidt) diagram
setEPS()

#postscript("./doc/artwork/Fig.eps", width=12, height=10)

pdf("./doc/artwork/Schmidt.pdf", width=10, height=10)

par(ps=12, cex=1, cex.main=2, cex.axis=1, cex.lab=1)

net()
qpoint(deg(bearing), 90-deg(plunge), col=4, pch=16, cex=1)
circular::rose.diag(bearing.circ, bins=12, col="darkgrey", prop=1.5, add=TRUE)
arrows.circular(mean(bearing.circ))
#lowplane(360, 3.908544, col=2, UP=TRUE, PLOT=TRUE)

dev.off()
```

### Randomness testing in three-dimensional orientation data

```{r}
# Angles are in radians, not f+$*@#in' degrees!!!
plunge <- rad(X.circ$marks.plunge)
bearing <- rad(X.circ$marks.bearing)

# step 1) express each direction in the sample as a set of three direction cosines. If the data are in the form of trend and plunge, the formulae are:
## l=cos(plunge)*cos(trend)
l <- c(cos(plunge)*cos(bearing))
## m=cos(plunge)*sin(trend)
m <- c(cos(plunge)*sin(bearing))
## n=sin(plunge)
n <- c(sin(plunge))
# step 2) and 3) compute the products of each direction-cosine set. sum the products over the whole data sample to give the orientation tensor
b <- matrix(c(
  sum(l^2),sum(l*m),sum(l*n),
  sum(m*l),sum(m^2),sum(m*n),
  sum(n*l),sum(n*m),sum(n^2)),
  ncol=3, byrow=TRUE)
# step 4) compute the eigenvectors and eigenvalues of this matrix
## the eigenvectors (v1,v2,v3) of this matrix correspond to three orthogonal 'principal axes' of the data sample. Axis v1 is an estimate of the mean direction of the data; v3 is the pole to any girdle present in the data [@Watson1966]
eigenvectors <- eigen(b)$vectors
dimnames(eigenvectors) <- list(c("v1","v2","v3"),c("l","m","n"))
## The eigenvalues (λ1,λ2,λ3), associated with (v1,v2,v3) respectively, describe the shape of the data sample [@Watson1966; @Mark1974; @Woodcock1973; @Woodcock1977]
eigenvalues <- eigen(b)$values
names(eigenvalues) <- c("λ1","λ2","λ3")
```

```{r, message=FALSE, warning=FALSE, results='hide'}
## equal area (schmidt) diagram
v1.plunge <- deg(asin(eigenvectors["v1","n"]))
v1.bearing <- deg(acos(eigenvectors["v1","l"]/cos(asin(eigenvectors["v1","n"]))))
v2.plunge <- deg(asin(eigenvectors["v2","n"]))
v2.bearing <- deg(acos(eigenvectors["v2","l"]/cos(asin(eigenvectors["v2","n"]))))
v3.plunge <- deg(asin(eigenvectors["v3","n"]))
v3.bearing <- deg(acos(eigenvectors["v3","l"]/cos(asin(eigenvectors["v3","n"]))))

net()
# lowplane(360, 3.908544, col="red", UP=TRUE, PLOT=TRUE)
qpoint(deg(bearing), 90-deg(plunge), col=4, pch=16, cex=1)
qpoint(deg(v1.bearing), 90-deg(v1.plunge), col=2, pch=10, cex=2, lab="v1")
qpoint(deg(v2.bearing), 90-deg(v2.plunge), col=2, pch=10, cex=2, lab="v2")
qpoint(deg(v3.bearing), 90-deg(v3.plunge), col=2, pch=10, cex=2, lab="v3")
```

```{r}
# step 5) compute the eigenvalue ratio, particularly the ratio of the largest to smallest eigenvalue (S1/S3=λ1/λ3)
## since λ1+λ2+λ3=N then normalized eigenvalues Sj=λj/N are related by S1+S2+S3=1. A data cluster has S1>S2~=S3 and a data girdle has S1~=S2>S3
s1 <- eigenvalues["λ1"]/length(plunge)
s2 <- eigenvalues["λ2"]/length(plunge)
s3 <- eigenvalues["λ3"]/length(plunge)
s1_s2 <- s1/s2
s2_s3 <- s2/s3
s1_s3 <- s1/s3
```

```{r}
## eigenvalues (Woodcock) ratio graph

plot(log(s2_s3),log(s1_s2),pch=15,xlim=c(0,3),ylim=c(0,3),xlab="ln S2/S3",ylab="ln S1/S2", main="")
## shape parameter K=ln(S1/S2)/ln(S2/S3) ranges from 0 (uniaxial girdles) to infinite (uniaxial clusters)
abline(h=0,col="grey",lty=2)
abline(v=0,col="grey",lty=2)
abline(0,1,col="grey",lty=2)
abline(0,2,col="grey",lty=2)
abline(0,.5,col="grey",lty=2)
abline(0,5,col="grey",lty=2)
abline(0,.2,col="grey",lty=2)
text(3,3,label="K=1",col="grey",lwd=.02)
text(3,.1,label="K=0",col="grey",lwd=.02)
text(.1,3,label="K=∞",col="grey",lwd=.02)
text(3,.75,label="K=0.2",col="grey",lwd=.02)
text(3,1.65,label="K=0.5",col="grey",lwd=.02)
text(.7,3,label="K=5",col="grey",lwd=.02)
text(1.6,3,label="K=2",col="grey",lwd=.02)
## complementary to the shape parameter is the parameter C=ln(S1/S3) which expresses the 'strength' of the preferred orientation in the data sample [@Woodcock1977]. Strongly organized samples have larger C and plot further from the origin of the ratio graph. A perfect uniform distribution of data would have C=0 and would plot at the origin with S1=S2=S3=1/3. Random samples will also plot close to the origin and have low C values, suggesting that this parameter might be used as a test statistic for randomness
abline(3,-1,col="grey",lty=1)
text(1.7,1.5,label="C=3",col="grey",lwd=.02)
abline(log(s1_s3),-1,col="red",lty=1)
text(1.4,1,label="ln S1/S3",col="red",lwd=.02)
##
PlotPTsmooth(deg(bearing),deg(plunge),x=log(s2_s3),y=log(s1_s2),siz=0.2,IMAGE=TRUE,CONT=TRUE,add=TRUE)
```

```{r}
# step 6) refer to Fig.9 or Table 1 and read off the critical S1/S3 value for the appropriate sample size and required confidence level. Note that for a small sample size a 95% confidence level is appropriate
## for N=300 the S1/S3 critical value at 95% confidence level is: 1.34
## for N=282?
# step 7) if the computed value is _less_ than the test value, the data sample cannot be distinguished from a random sample at this confidence level
s1s3_test <- read.csv(url("https://raw.githubusercontent.com/dncgst/woodcock/master/vignettes/Woodcock%26Naylor1983_fig9.csv"), header=TRUE, sep=",", skip=0)
str(s1s3_test)
C <- c(s1_s3,263)

plot(x=s1s3_test$N, y=log(s1s3_test$X90.), type="b", col="grey", lty=2, xlim=c(0,300), xlab="Obs.", ylab="ln C", main="b")
lines(x=s1s3_test$N, y=log(s1s3_test$X95.), type="b", col="grey", lty=2, add=TRUE)
lines(x=s1s3_test$N, y=log(s1s3_test$X97.5.), type="b", col="grey", lty=2, add=TRUE)
lines(x=s1s3_test$N, y=log(s1s3_test$X99.), type="b", col="grey", lty=2, add=TRUE)
points(x=C[2], y=log(C[1]), pch=21, bg="red", col="red", cex=2, add=TRUE)
text(10,1, label="RANDOM",col="grey", lwd=.02)
text(25,2, label="SIGNIFICANT",col="grey",lwd=.02)
text(183,2, label="",col="black",lwd=.02)
```

The C value (`r log(s1_s3)`) is higher than the critical S1/S3 test value (1.44) for N=300 at 99% confidence, thus the data sample reject the hypothesis of randomness in favor of a strong organized sample. The K value (`r log(s1_s2)/log(s2_s3)`) plots the data sample close to K=0, indicating uniaxial girdles (planar fabric).

### Benn diagram

```{r}
# isotropy index
is <- s3/s1
# elongation index
el <- 1-(s2/s1)
# 
bs <- 1-(is+el)

# Benn diagram
triax.plot(matrix(c(bs,el,is),ncol=3,byrow=2), show.grid=TRUE, pch=16, cex.ticks=.8, cex.axis=1, axis.labels=c("","Elongation Index\n EL=1-(S2/S1)","Isotropy Index\n IS=S3/S1"), tick.labels=list(l=seq(0.1,0.9,by=0.1), r=seq(0.1,0.9,by=0.1), b=NULL), main="Benn diagram")
triax.points(matrix(c(bs,el,is),ncol=3,byrow=2), pch=16, cex=2, col="red")
text(0.5,0.9,label="isotropic")
text(-0.1,0,label="planar")
text(1.1,0,label="linear")
```

#### Fig.5

```{r}
# Fig. Rose diagram (a). Equal area Schimdt diagram (b). Woodcock (c) and Benn (d) diagrams.
setEPS()

pdf("./doc/manuscript/artwork/Fig:fabric_.pdf", width=30, height=10)

#postscript("./doc/artwork/Fig.eps", width=12, height=10)

par(mfcol=c(1,3), mar= c(6,6,6,6), ps=18, cex=1, cex.main=2, cex.axis=1, cex.lab=1)

## rose diagram
heR.Misc::rose(X.circ$marks.bearing, bins=16, col="#e41a1c", rings=TRUE, labels=TRUE, rscale=2)

# Benn diagram
# triax.plot(matrix(c(bs,el,is),ncol=3,byrow=2), show.grid=TRUE, pch=16, cex.ticks=.8, cex.axis=1, axis.labels=c("","Elongation Index\n EL=1-(S2/S1)","Isotropy Index\n IS=S3/S1"), tick.labels=list(l=seq(0.1,0.9,by=0.1), r=seq(0.1,0.9,by=0.1), b=NULL), main="a")
# triax.points(matrix(c(bs,el,is),ncol=3,byrow=2), pch=16, cex=4, col="#595959ff")
# text(0.5,0.9,label="isotropic")
# text(-0.1,0,label="planar")
# text(1.1,0,label="linear")

## equal area (Schmidt) diagram
net()
# lowplane(360, 3.908544, col="red", UP=TRUE, PLOT=TRUE)
qpoint(deg(bearing), 90-deg(plunge), col="#595959ff", pch=16, cex=1.5)
qpoint(deg(v1.bearing), 90-deg(v1.plunge), col=2, pch=13, cex=3, lab="v1")
qpoint(deg(v2.bearing), 90-deg(v2.plunge), col=2, pch=13, cex=3, lab="v2")
qpoint(deg(v3.bearing), 90-deg(v3.plunge), col=2, pch=13, cex=3, lab="v3")
#circular::rose.diag(bearing.circ, bins=16, col="#e41a1c", prop=1.5, add=TRUE)
#arrows.circular(mean(bearing.circ))

## eigenvalues (Woodcock) ratio graph
plot(log(s2_s3),log(s1_s2),pch=15,xlim=c(0,3),ylim=c(0,3),xlab="ln S2/S3",ylab="ln S1/S2", main="b")
abline(h=0,col="grey",lty=2)
abline(v=0,col="grey",lty=2)
abline(0,1,col="grey",lty=2)
abline(0,2,col="grey",lty=2)
abline(0,.5,col="grey",lty=2)
abline(0,5,col="grey",lty=2)
abline(0,.2,col="grey",lty=2)
text(2.95,2.8,label="K=1",col="grey",lwd=.02)
text(2.95,.12,label="K=0",col="grey",lwd=.02)
text(.2,3,label=expression(paste("K=",infinity)),col="grey",lwd=.02)
text(2.95,.75,label="K=0.2",col="grey",lwd=.02)
text(2.95,1.65,label="K=0.5",col="grey",lwd=.02)
text(.8,3,label="K=5",col="grey",lwd=.02)
text(1.7,3,label="K=2",col="grey",lwd=.02)
abline(3,-1,col="grey",lty=1)
text(1.7,1.5,label="C=3",col="grey",lwd=.02)
text(1.45,1.65,label="WEAK <--> STRONG",col="gray",lwd=.03,srt=45,pos=1)
text(2.55,2.63,label="<-- CLUSTERS | GIRDLES -->",col="gray",lwd=.03,srt=-45,pos=1)
#abline(log(s1_s3),-1,col="red",lty=1)
#text(1.4,1,label="ln S1/S3",col="red",lwd=.02)
PlotPTsmooth(deg(bearing),deg(plunge),x=log(s2_s3),y=log(s1_s2),siz=0.2,IMAGE=TRUE,CONT=TRUE,add=TRUE)

dev.off()
```

## References